---
title: 'DATA 612 Project 1 | Global Baseline Predictors and RMSE'
author: "Joseph E. Garcia"
date: "June 8, 2019"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
```
## Details

*Briefly describe the recommender system that you're going to build out from a business
perspective, e.g. "This system recommends data science books to readers."

*Find a dataset, or build out your own toy dataset. As a minimum requirement for complexity,
please include numeric ratings for at least five users, across at least five items, with some missing
data.

*Load your data into (for example) an R or pandas dataframe, a Python dictionary or list of lists, (or
another data structure of your choosing). From there, create a user-item matrix.

*If you choose to work with a large dataset, you're encouraged to also create a small, relatively
dense "user-item" matrix as a subset so that you can hand-verify your calculations.

*Break your ratings into separate training and test datasets.

*Using your training data, calculate the raw average (mean) rating for every user-item combination.

*Calculate the RMSE for raw average for both your training data and your test data.

*Using your training data, calculate the bias for each user and each item.

*From the raw average, and the appropriate user and item biases, calculate the baseline predictors
for every user-item combination.

*Calculate the RMSE for the baseline predictors for both your training data and your test data.

*Summarize your results.

## Data Prep

```{r}
set.seed(643)
n <- 15
percent <- .3
probs <- c(1-percent, percent)
users <- paste("User", LETTERS[1:n], sep="")
items <- paste("Item", LETTERS[1:n], sep="")

toyDF <- as.data.frame(replicate(n, floor(runif(n, 1,6))))

toyDF <- as.data.frame(
    lapply(toyDF, function(df) df[sample(c(TRUE, NA), prob = probs, size = length(df), replace = TRUE) ]), 
    row.names = users, 
    col.names = items)
knitr::kable(toyDF)
```

## Data Visualization

```{r}
heatmap(as.matrix(toyDF), Rowv=NA, Colv=NA, col = terrain.colors(256))
```

## Separate Data 

```{r}
train_row <- sample(n, size = floor(2/3 * n), replace=F)
train <- toyDF[train_row, ]
test <- toyDF[-train_row, ]
```

## Calculations


```{r}
train_raw_avg <- mean(as.matrix(train), na.rm=TRUE)
train_raw_avg
```


```{r}
RMSE <- function(observed, predicted) {
   
    sqrt(mean((observed - predicted)^2, na.rm=TRUE))
}
trainRMSE <- RMSE(as.matrix(train), train_raw_avg)
trainRMSE
testRMSE <- RMSE(as.matrix(test), train_raw_avg)
testRMSE
```


```{r}
user_bias_train <- rowMeans(as.matrix(train), na.rm = T) - train_raw_avg
user_bias_test <- rowMeans(as.matrix(test), na.rm = T) - train_raw_avg
item_bias_train <- colMeans(as.matrix(train), na.rm = T) - train_raw_avg
user_bias_train
user_bias_test
item_bias_train
```



```{r}
baseline_predictor <- function(rawAvg, userBias, itemBias) {
   
    userlist <- names(userBias)
    itemlist <- names(itemBias)
    df <- data.frame()
    
    for (i in userBias) {
        UserPred <- rawAvg + i + itemBias
        df <- rbind(df, UserPred)
    }
    
 
    df[df > 5] <- 5
    df[df < 1] <- 1
    
    row.names(df) <- userlist
    names(df) <- itemlist
    df
}
trainPredictions <- baseline_predictor(train_raw_avg, user_bias_train, item_bias_train)
heatmap(as.matrix(trainPredictions), Rowv=NA, Colv=NA, scale = 'none', col = topo.colors(256))
knitr::kable(trainPredictions)
testPredictions <- baseline_predictor(train_raw_avg, user_bias_test, item_bias_train)
knitr::kable(testPredictions)
```



```{r}
trainPredictionRSME <- RMSE(as.matrix(train), as.matrix(trainPredictions))
trainPredictionRSME
testPredictionRSME <- RMSE(as.matrix(test), as.matrix(testPredictions))
testPredictionRSME
```

## Summary


```{r}
trainVec <- c(trainRMSE,
              trainPredictionRSME,
              (1-trainPredictionRSME/trainRMSE)*100)
testVec <- c(testRMSE, 
             testPredictionRSME, 
             (1-testPredictionRSME/testRMSE)*100)
summary <- data.frame(trainVec, testVec)
names(summary) <- c("train", "test")
row.names(summary) <- c("Raw Average RMSE", "Simple Predictor RMSE", "Percent Improvement")
knitr::kable(summary)
```


