---
title: "Data 612 - Project#3 Matrix Factorization Methods"
author: "Joseph E. Garcia"
date: "June 20, 2019"
output: html_document
---
The goal of this assignment is give you practice working with Matrix Factorization techniques.
Your task is implement a matrix factorization method-such as singular value decomposition (SVD) or
Alternating Least Squares (ALS)-in the context of a recommender system.
You may approach this assignment in a number of ways. You are welcome to start with an existing
recommender system written by yourself or someone else. Remember as always to cite your sources,
so that you can be graded on what you added, not what you found.
SVD can be thought of as a pre-processing step for feature engineering. You might easily start with
thousands or millions of items, and use SVD to create a much smaller set of "k" items (e.g. 20 or 70).

# Data input

## Loading libraries/dataset
```{r results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(recommenderlab)
library(reshape2)
library(knitr)
```

```{r}
movies = read.csv("https://raw.githubusercontent.com/jgarcia71/Data-612-Projects-and-Assignments-Summer-2019/master/Project%20%233/movies.csv", 
                   header = TRUE, sep = ",", stringsAsFactors = FALSE)
ratings = read.csv("https://raw.githubusercontent.com/jgarcia71/Data-612-Projects-and-Assignments-Summer-2019/master/Project%20%233/ratings.csv", 
                    header = TRUE, sep =",", stringsAsFactors = FALSE)
kable(head(ratings))
kable(head(movies))
```

  
```{r}
movRate = merge(movies, ratings, by = "movieId")
new = subset(movRate, select = c("title", "userId", "rating"))
new = unique(new)
kable(head(new))
matrix = acast(new, userId~title, value.var="rating", fun=sum)
```


```{r}
premat = as(matrix, "realRatingMatrix")
data = premat[sample(671,200), sample(9064, 3000)]
data
```

# Data Exploration
.
```{r}
# Ratings
qplot(new$rating, geom="histogram", main = "Histogram of Ratings", xlab = "Rating Scores", binwidth = 0.5, fill=I("yellow"))
```

```{r}
# Ratings(Movie)
new2 = new %>% group_by(title) %>%
  summarise(count = mean(rating))
qplot(new2$count, geom="histogram", main = "Histogram of Movie Ratings", xlab = "Average Rating Scores Per Movie", binwidth = 0.5, fill=I("blue"))
```


```{r}
# Ratings(User)
new3 = new %>% group_by(userId) %>%
  summarise(count = mean(rating))
qplot(new3$count, geom="histogram", main = "Histogram of User Ratings", xlab = "Average Rating Scores Per User", binwidth = 0.5, fill=I("red"))
```


```{r}
#Evaluation
evaluation = evaluationScheme(data, method="split", train=0.8, given=10, goodRating=3.5)
ev_train = getData(evaluation, "train")
ev_known = getData(evaluation, "known")
ev_unknown = getData(evaluation, "unknown")
```

# SVD

```{r}
example = as.matrix(data.frame(c(1,3,4,0), c(1,2,4,0), c(0,0,0,5)))
example
svd(example)
```


```{r}
svd_train = Recommender(ev_train, "SVD")
svd_preds = predict(svd_train, ev_known, type = "ratings")
getRatingMatrix(svd_preds[c(1,9,17,25,33),1:5])
```


# Comparison


```{r}

ubcf_train = Recommender(ev_train, "UBCF")
ubcf_preds = predict(ubcf_train, ev_known, type = "ratings")

pop_train = Recommender(ev_train, "POPULAR")
pop_preds = predict(pop_train, ev_known, type = "ratings")
accuracy = rbind(
  SVD = calcPredictionAccuracy(svd_preds, ev_unknown),
  UBCF = calcPredictionAccuracy(ubcf_preds, ev_unknown),
  POPULAR = calcPredictionAccuracy(pop_preds, ev_unknown)
  )
kable(as.data.frame(accuracy))
```


```{r}
eval_sets = evaluationScheme(data = data, method = "cross-validation", k = 4, given = 10, goodRating = 3.5)
mult_models = list(
  UBCF = list(name = "UBCF", param = list(method = "pearson")),
  Popular = list(name = "POPULAR", param = NULL),
  SVD = list(name = "SVD", param = NULL)
)

models = evaluate(eval_sets, mult_models, n= c(1, 5, seq(10, 100, 10)))

plot(models, annotate = T, legend="topleft")
plot(models, "prec/rec", annotate = F, main="Precision/Recall", legend="topright")
```


